Supplemental data to ``\emph{k}-mer-based GWAS enhances causal variant and candidate gene discovery in soybean''

### 2023-02-28 additional_files/additional_file_1.tex
Adjust the text to remove any mention of vg

We conducted two different GWAS analyses on seed coat color. The first GWAS
included five different phenotypes observed in the dataset (Table
\ref{SEED.COAT.COLOR_gwas_table}) and resulted in the detection of strong
signals overlapping the \emph{I} locus by all four approaches (Figure
\ref{seed_coat_color_all_manhattan}). All approaches but vg also detected
signals at the \emph{G} locus. In addition, the \emph{k}-mer approach also
detected weak signals overlapping the \emph{T} and \emph{O} loci. However, the
\emph{T} and \emph{O} loci are likely chance associations given the low \textit{p}-value
associated with the \emph{T} locus and the proximity of the \emph{O} locus to
the \emph{I} locus. Detailed results of this first GWAS and the \textit{I}
locus can be viewed in figures \ref{seed_coat_color_all_manhattan} and
\ref{seed_coat_color_all_I_signal}.

@@

We conducted three different GWAS analyses on hilum color. The first GWAS
included seven colors (Table \ref{HILUM.COLOR_gwas_table}) and detected signals
at the \emph{I}, \emph{T} and \emph{R} loci using all four approaches except vg
at the \emph{R} locus (Figure \ref{hilum_color_all_manhattan}). No signals
were detected at the \emph{W1} and \emph{O} loci studied by
\cite{bandillo2017}. The detailed results for the \textit{I} locus can be viewed
in figure \ref{hilum_color_all_I_signal} whereas those for the \textit{T} locus
can be viewed in figures \ref{hilum_color_all_T_signal}, \ref{hilum_color_all_T_gene}
and \ref{hilum_color_all_T_kmers}.

@@

We performed three different GWAS analyses on seed coat luster (Table
\ref{SEED.COAT.LUSTER_gwas_table}). Since all three analyses yielded similar results
(Figure \ref{seed_coat_luster_all_manhattan},
\ref{seed_coat_luster_nointermediate_manhattan},
\ref{seed_coat_luster_dullshiny_manhattan}), we chose to focus on the
GWAS contrasting the dull and shiny phenotypes as it yielded the most
significant signals. In addition, it also provided results that were most consistent with the
analysis previously done by \cite{bandillo2017}. The analysis detected
signals at the \emph{B} locus as the sole classical locus using all four
methods (Figure \ref{seed_coat_luster_dullshiny_manhattan}). Apparent signals
on chromosomes Gm20 and Gm09 (which may correspond to the B? locus reported by
\cite{bandillo2017}) were also detected, but analysis of the LD patterns
between \emph{k}-mers suggested that these signals were most likely in LD with
the \emph{B} locus (Figure \ref{seed_coat_luster_dullshiny_ld}). All methods
also detected signals overlapping the \emph{Hps} locus as defined by
\cite{bandillo2017}, however results shown by \cite{gijzen2003} and our results
suggest that \emph{Hps} and \emph{B} are the same locus. Using a sequencing depth-based segmentation analysis
similar to the one used for the \emph{Ps} locus, we identified a 31-kb region
spanning positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited
variation in copy number across accessions and overlapped four genes annotated
as containing HPS domains (Figures \ref{seed_coat_luster_dullshiny_B_signal},
\ref{seed_coat_luster_dullshiny_B_gene}). The top 5\% / 1\% significant
associations of all approaches other than vg overlapped this CNV region and the
most significantly associated \emph{k}-mer was located within the CNV region
(Figures \ref{seed_coat_luster_dullshiny_B_signal},
\ref{seed_coat_luster_dullshiny_B_gene}). However we were not able to link any
of the variants found to the presence of copy number variation.

@@

Similarly to seed oil content, we conducted a GWAS on seed protein content
using data obtained from the GRIN database. As is generally the case in
soybean given the negative correlation between seed oil and protein content,
the two major QTL considered above and reported by \cite{bandillo2015} for seed
oil content are also major QTL for seed protein content. We only identified a signal
at the locus located on chromosome Gm20, and only using the genotypes
called by vg (Figure \ref{protein_manhattan}). However, the SV discovered as
significantly associated was located about 700 kb from the gene cloned by
\cite{fliege2022} for that locus (Glyma.20g085100). It is therefore unlikely that this SV
represents the causal variation at that locus.

>>>>

We conducted two different GWAS analyses on seed coat color. The first GWAS
included five different phenotypes observed in the dataset (Table
\ref{SEED.COAT.COLOR_gwas_table}) and resulted in the detection of strong
signals overlapping the \emph{I} locus by all four approaches (Figure
\ref{seed_coat_color_all_manhattan}). All approaches also detected
signals at the \emph{G} locus. In addition, the \emph{k}-mer approach also
detected weak signals overlapping the \emph{T} and \emph{O} loci. However, the
\emph{T} and \emph{O} loci are likely chance associations given the low \textit{p}-value
associated with the \emph{T} locus and the proximity of the \emph{O} locus to
the \emph{I} locus. Detailed results of this first GWAS and the \textit{I}
locus can be viewed in figures \ref{seed_coat_color_all_manhattan} and
\ref{seed_coat_color_all_I_signal}.

@@

We conducted three different GWAS analyses on hilum color. The first GWAS
included seven colors (Table \ref{HILUM.COLOR_gwas_table}) and detected signals
at the \emph{I}, \emph{T} and \emph{R} loci using all four approaches
(Figure \ref{hilum_color_all_manhattan}). No signals
were detected at the \emph{W1} and \emph{O} loci studied by
\cite{bandillo2017}. The detailed results for the \textit{I} locus can be viewed
in figure \ref{hilum_color_all_I_signal} whereas those for the \textit{T} locus
can be viewed in figures \ref{hilum_color_all_T_signal}, \ref{hilum_color_all_T_gene}
and \ref{hilum_color_all_T_kmers}.

@@

We performed three different GWAS analyses on seed coat luster (Table
\ref{SEED.COAT.LUSTER_gwas_table}). Since all three analyses yielded similar results
(Figure \ref{seed_coat_luster_all_manhattan},
\ref{seed_coat_luster_nointermediate_manhattan},
\ref{seed_coat_luster_dullshiny_manhattan}), we chose to focus on the
GWAS contrasting the dull and shiny phenotypes as it yielded the most
significant signals. In addition, it also provided results that were most consistent with the
analysis previously done by \cite{bandillo2017}. The analysis detected
signals at the \emph{B} locus as the sole classical locus using all four
methods (Figure \ref{seed_coat_luster_dullshiny_manhattan}). Apparent signals
on chromosomes Gm20 and Gm09 (which may correspond to the B? locus reported by
\cite{bandillo2017}) were also detected, but analysis of the LD patterns
between \emph{k}-mers suggested that these signals were most likely in LD with
the \emph{B} locus (Figure \ref{seed_coat_luster_dullshiny_ld}). All methods
also detected signals overlapping the \emph{Hps} locus as defined by
\cite{bandillo2017}, however results shown by \cite{gijzen2003} and our results
suggest that \emph{Hps} and \emph{B} are the same locus. Using a sequencing depth-based segmentation analysis
similar to the one used for the \emph{Ps} locus, we identified a 31-kb region
spanning positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited
variation in copy number across accessions and overlapped four genes annotated
as containing HPS domains (Figures \ref{seed_coat_luster_dullshiny_B_signal},
\ref{seed_coat_luster_dullshiny_B_gene}). The top 5\% / 1\% significant
associations of all approaches overlapped this CNV region and the
most significantly associated \emph{k}-mer was located within the CNV region
(Figures \ref{seed_coat_luster_dullshiny_B_signal},
\ref{seed_coat_luster_dullshiny_B_gene}). However we were not able to link any
of the variants found to the presence of copy number variation.

@@

Similarly to seed oil content, we conducted a GWAS on seed protein content
using data obtained from the GRIN database. As is generally the case in
soybean given the negative correlation between seed oil and protein content,
the two major QTL considered above and reported by \cite{bandillo2015} for seed
oil content are also major QTL for seed protein content. However, none of the
approaches detected a signal at this locus.

### 2023-02-28 additional_files/main_text.tex
Adjust the text to remove any mention of vg

To help with the interpretation of results and the identification of candidate
genes, we defined regions of highly significant associations within each signal
by delimiting a range of coordinates containing the top 5\% most significant
associations (Paragraph and vg) or 1\% most significant associations (Platypus
and \emph{k}-mers) found within the signal. In cases where such a fraction
represented fewer than two markers or \emph{k}-mers, the highly significant
region was simply delimited by the two most significant associations.

@@

We analyzed a total of 13 traits using three different GWAS approaches, one
based on SNP and indel genotypes (Platypus), one based on SV genotypes (Paragraph),
and one based on the presence/absence of \textit{k}-mers.
Because two or three different GWAS analyses were conducted for some of the traits, a
total of 22 GWAS analyses were performed using each of the four approaches.
Given the large volume of data that this represents, we will focus here on the
most noteworthy results. Readers interested in detailed results for loci not
discussed here are referred to Supplemental data. In particular, table \ref{signals_table}
provides a summary of the \textit{p}-values obtained by each approach at each
of the loci considered in this study.

@@

% Table 1 with results on all loci with cloned genes
\begin{landscape}
	\begin{table}
		\caption[Performance of four GWAS approaches at detecting causal genes/variants]%
		{Performance of four GWAS approaches at detecting causal genes or variants at cloned loci.}
		\label{loci-table}

@@

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny and light
tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we
detected signals in this region using all four approaches, but only Platypus
and the \textit{k}-mers identified a signal that overlapped Glyma.03g258700
(Figures \ref{pubescence-color-main-figure}a-b,
\ref{pubescence_color_nogray_manhattan}, \ref{pubescence_color_nogray_Td_signal}).
While the most significant \textit{k}-mer identified the causal variant documented
at this locus \citep{yan2020}, it was the fourth most significantly associated variant
for the Platypus genotype calls (Figures
\ref{pubescence-color-main-figure}b-d, \ref{pubescence_color_nogray_Td_gene}).
Interestingly, some of the accessions bearing the haplotype associated with tawny pubescence
displayed light tawny pubescence, suggesting that there may be more than one causal
variant at this locus (Figure \ref{pubescence-color-main-figure}e).

@@

The \textit{T} locus mentioned above for pubescence color is also a determining
locus for hilum color, and we detected signals overlapping the corresponding gene
using all approaches but vg (Figures \ref{hilum_color_all_manhattan},
\ref{hilum_color_all_T_signal}). The regions defined by the top 5\% (Paragraph)
and top 1\% (Platypus and \textit{k}-mers) variants/\textit{k}-mers included
the Glyma.06g202300 gene associated with this locus. Similarly to what we
observed for the analysis of the \textit{T} locus based on pubescence color, we
did identify the documented causal variant at this locus using the Platypus and
\textit{k}-mer analyses, although these variants were not the most
significantly associated overall (Figures \ref{hilum_color_all_T_gene},
\ref{hilum_color_all_T_kmers}).

@@

We used corrected dry weight (CDW) as described by \cite{deronne2022} as a
measure of horizontal resistance to \emph{P. sojae}. We found a strong signal
overlapping the genomic region reported by these authors on Gm15 using all four GWAS
approaches (Figure \ref{corrected_dry_weight_manhattan}). \cite{deronne2022}
proposed Glyma.15g217100 as a candidate gene for explaining phenotypic
variation at that locus based on a combination of functional annotation and
contrasting gene expression in resistant and susceptible lines four days after
inoculation. Although the signals found by our analyses did overlap
Glyma.15g217100, we found that this gene was located at the distal end of the signal
in a region with markers or \emph{k}-mers showing a weaker
association (Figure \ref{corrected_dry_weight_cdwGm15_signal}). The
large size of the observed signal means that the gene associated with the phenotype could
be located almost anywhere in this region, however, we identified a particularly
interesting region defined by the top 1\% associations in the Platypus
analysis. This region overlaps only three annotated genes and also contains the
most significant \emph{k}-mer (Figure
\ref{corrected_dry_weight_cdwGm15_signal}). One of the genes in this region is
a calcium ion binding protein with an EF-hand domain (Glyma.15g217700).
Given the important role played by calcium signaling in defense
responses \citep{zhang2014}, we suggest Glyma.15g217700 as an interesting
candidate gene for explaining resistance to \emph{P. sojae} in this population.

>>>>

To help with the interpretation of results and the identification of candidate
genes, we defined regions of highly significant associations within each signal
by delimiting a range of coordinates containing the top 5\% most significant
associations (Paragraph) or 1\% most significant associations (Platypus
and \emph{k}-mers) found within the signal. In cases where such a fraction
represented fewer than two markers or \emph{k}-mers, the highly significant
region was simply delimited by the two most significant associations.

@@

We analyzed a total of 13 traits using three different GWAS approaches, one
based on SNP and indel genotypes (Platypus), one based on SV genotypes (Paragraph),
and one based on the presence/absence of \textit{k}-mers.
Because two or three different GWAS analyses were conducted for some of the traits, a
total of 22 GWAS analyses were performed using each of the all approaches.
Given the large volume of data that this represents, we will focus here on the
most noteworthy results. Readers interested in detailed results for loci not
discussed here are referred to Supplemental data. In particular, table \ref{signals_table}
provides a summary of the \textit{p}-values obtained by each approach at each
of the loci considered in this study.

@@

% Table 1 with results on all loci with cloned genes
\begin{landscape}
	\begin{table}
		\caption[Performance of three GWAS approaches at detecting causal genes/variants]%
		{Performance of three GWAS approaches at detecting causal genes or variants at cloned loci.}
		\label{loci-table}

@@

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny and light
tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we
detected signals in this region using all three approaches, but only Platypus
and the \textit{k}-mers identified a signal that overlapped Glyma.03g258700
(Figures \ref{pubescence-color-main-figure}a-b,
\ref{pubescence_color_nogray_manhattan}, \ref{pubescence_color_nogray_Td_signal}).
While the most significant \textit{k}-mer identified the causal variant documented
at this locus \citep{yan2020}, it was the fourth most significantly associated variant
for the Platypus genotype calls (Figures
\ref{pubescence-color-main-figure}b-d, \ref{pubescence_color_nogray_Td_gene}).
Interestingly, some of the accessions bearing the haplotype associated with tawny pubescence
displayed light tawny pubescence, suggesting that there may be more than one causal
variant at this locus (Figure \ref{pubescence-color-main-figure}e).

@@

The \textit{T} locus mentioned above for pubescence color is also a determining
locus for hilum color. We detected signals overlapping the corresponding gene
using all approaches (Figures \ref{hilum_color_all_manhattan},
\ref{hilum_color_all_T_signal}). The regions defined by the top 5\% (Paragraph)
and top 1\% (Platypus and \textit{k}-mers) variants/\textit{k}-mers included
the Glyma.06g202300 gene associated with this locus. Similarly to what we
observed for the analysis of the \textit{T} locus based on pubescence color, we
did identify the documented causal variant at this locus using the Platypus and
\textit{k}-mer analyses, although these variants were not the most
significantly associated overall (Figures \ref{hilum_color_all_T_gene},
\ref{hilum_color_all_T_kmers}).

@@

We used corrected dry weight (CDW) as described by \cite{deronne2022} as a
measure of horizontal resistance to \emph{P. sojae}. We found a strong signal
overlapping the genomic region reported by these authors on Gm15 using all all GWAS
approaches (Figure \ref{corrected_dry_weight_manhattan}). \cite{deronne2022}
proposed Glyma.15g217100 as a candidate gene for explaining phenotypic
variation at that locus based on a combination of functional annotation and
contrasting gene expression in resistant and susceptible lines four days after
inoculation. Although the signals found by our analyses did overlap
Glyma.15g217100, we found that this gene was located at the distal end of the signal
in a region with markers or \emph{k}-mers showing a weaker
association (Figure \ref{corrected_dry_weight_cdwGm15_signal}). The
large size of the observed signal means that the gene associated with the phenotype could
be located almost anywhere in this region, however, we identified a particularly
interesting region defined by the top 1\% associations in the Platypus
analysis. This region overlaps only three annotated genes and also contains the
most significant \emph{k}-mer (Figure
\ref{corrected_dry_weight_cdwGm15_signal}). One of the genes in this region is
a calcium ion binding protein with an EF-hand domain (Glyma.15g217700).
Given the important role played by calcium signaling in defense
responses \citep{zhang2014}, we suggest Glyma.15g217700 as an interesting
candidate gene for explaining resistance to \emph{P. sojae} in this population.
